### 2025-04-03 scripts/Paper-analysis.Rmd
updating analysis

```
```{r}
# Sliding window for GC content
p.lines.GC = ggplot(d.reduced.median.plot, aes(x, GC/5000 * 100))+
  geom_hline(yintercept = 0, linetype='dashed')+
  theme_bw()+
  geom_path(aes(group=PTU, alpha=normalized_n))+
  facet_wrap(~PTU, scales="free_x", nrow=3)+
  ylab("Palindrome depletion per kb (5kb sliding window)")+
  labs(alpha="% of plasmids within PTU")+
  theme(panel.grid=element_blank())+  
  scale_x_continuous(labels = function(x) x / 1000) +
  xlab("Position on plasmid (kb)")+
  theme(legend.position = c(0.8, 0.2))
ggsave(p.lines.GC, file='../results/GC-content.pdf',
       width=10, height=6)

@@

```{r}
# Start from d.reduced, which contains per-plasmid results with 5000bp sliding window and step size of 500bp
# Compare depletion in first 5kb with average for the rest of the plasmid
diff.leading.region.average = d.reduced %>% 
  group_by(plasmid, PTU) %>%
  mutate(diff.count=palindrome_count[x=="0"]-mean(palindrome_count),
         diff.density=diff_density[x=="0"]-mean(diff_density)) %>%
  filter(x=="0")

# Order the PTUs by mean values of depletion for better visual display
mean.values = diff.leading.region.average %>% group_by(PTU) %>%
  summarise(mean=mean(diff.density)*1000)
diff.leading.region.average$PTU = ordered(gsub("PTU-", "", diff.leading.region.average$PTU),
                                          levels=gsub("PTU-", "", mean.values$PTU[order(mean.values$mean)]))
p.leading.region.diff.density = ggplot(diff.leading.region.average, aes(as.factor(PTU), diff.density*1000))+ # depletion per-kb rather than per-bp
  ggbeeswarm::geom_quasirandom(alpha=0.8)+
  theme_bw()+
  theme(panel.grid = element_blank())+
  geom_hline(yintercept = 0)+
  xlab("")+
  theme(axis.text=element_text(colour="black", size=18))+
  ylab("Palindrome depletion in leading region (per kb)")+
  theme(axis.text.x=element_text(angle=45, hjust=1))+
  ylab("")
p.leading.region.diff.density
# Based on this, remove two outliers
p.leading.region.diff.density = p.leading.region.diff.density+
  scale_y_continuous(breaks=seq(-8, 4, 2), limits=c(-8,5))
ggsave(p.leading.region.diff.density, file='../results/6mer-deviation-leading-region-vs-rest-of-plasmid-density.pdf',
       width=5.5, height=6)
```
### Averaging over all plasmids

@@

```{r}
overall.median = d.reduced %>% filter(x<50000) %>%
  group_by(x) %>%
  summarise(y=median(diff_density),
            n=length(diff_density))
p.overall.average = ggplot(overall.median, aes(x, y*1000))+
  geom_line()+
  scale_y_continuous(breaks=seq(-10, 0, 2), limits = c(-9, 0))+
  theme_bw()+
  theme(panel.grid = element_blank())+
  geom_hline(yintercept = 0, linetype='dashed')+
  ylab("Palindrome depletion (per kb)")+
    scale_x_continuous(labels = function(x) x / 1000) +
  xlab("Position on plasmid (kb)")+
  theme(axis.text=element_text(colour="black"))
p.overall.average
# Save
ggsave(p.overall.average, file='../results/6mer-deviation-overall.pdf', 
       width=3, height=3)
```

>>>>

```
```{r}
# Sliding window for GC content
p.lines.GC = ggplot(d.reduced.median.plot, aes(x, GC/5000 * 100))+
  geom_hline(yintercept = 0, linetype='dashed')+
  theme_bw()+
  geom_path(aes(group=PTU, alpha=normalized_n))+
  facet_wrap(~PTU, scales="free_x", nrow=3)+
  ylab("GC content (5kb sliding window)")+
  labs(alpha="% of plasmids within PTU")+
  theme(panel.grid=element_blank())+  
  scale_x_continuous(labels = function(x) x / 1000) +
  xlab("Position on plasmid (kb)")+
  theme(legend.position = c(0.8, 0.2))
ggsave(p.lines.GC, file='../results/GC-content.pdf',
       width=10, height=6)

@@

```{r}
# Start from d.reduced, which contains per-plasmid results with 5000bp sliding window and step size of 500bp
# Compare depletion in first 5kb with average for the rest of the plasmid
diff.leading.region.average = d.reduced %>% 
  group_by(plasmid, PTU) %>%
  mutate(diff.count=palindrome_count[x=="0"]-mean(palindrome_count),
         diff.density=diff_density[x=="0"]-mean(diff_density),
         diff.GC=(GC_n[x=="0"]-mean(GC_n))/5000 * 100) %>%
  filter(x=="0")

# Order the PTUs by mean values of depletion for better visual display
mean.values = diff.leading.region.average %>% group_by(PTU) %>%
  summarise(mean=mean(diff.density)*1000)
diff.leading.region.average$PTU = ordered(gsub("PTU-", "", diff.leading.region.average$PTU),
                                          levels=gsub("PTU-", "", mean.values$PTU[order(mean.values$mean)]))
p.leading.region.diff.density = ggplot(diff.leading.region.average, aes(as.factor(PTU), diff.density*1000))+ # depletion per-kb rather than per-bp
  ggbeeswarm::geom_quasirandom(alpha=0.8)+
  theme_bw()+
  theme(panel.grid = element_blank())+
  geom_hline(yintercept = 0)+
  xlab("")+
  theme(axis.text=element_text(colour="black", size=18))+
  ylab("Palindrome depletion in leading region (per kb)")+
  theme(axis.text.x=element_text(angle=45, hjust=1))+
  ylab("")
p.leading.region.diff.density
# Based on this, remove two outliers
p.leading.region.diff.density = p.leading.region.diff.density+
  scale_y_continuous(breaks=seq(-8, 4, 2), limits=c(-8,5))
ggsave(p.leading.region.diff.density, file='../results/6mer-deviation-leading-region-vs-rest-of-plasmid-density.pdf',
       width=5.5, height=6)
```

Now we want to correlate the GC content difference of the first 5kb of the plasmid with the difference in palindrome density

```{r}
diff.PTU.averages = diff.leading.region.average %>% group_by(PTU) %>%
  summarise(mean.GC=mean(diff.GC),
            GC=mean(GC_n[x=="0"]/5000 * 100),
            mean.palindrome=mean(palindrome_density),
            mean.palindrome.diff=mean(diff.density) * 1000,
            n=length(diff.GC))
ggplot(diff.PTU.averages, aes(mean.GC, mean.palindrome.diff))+
  geom_hline(yintercept = 0, linetype='dashed')+
  ggrepel::geom_text_repel(aes(label=PTU))+
    geom_point(aes(size=n), shape=21, colour="black", fill="grey")+
  theme_bw()+
  theme(panel.grid = element_blank())+
  xlab("GC difference between first 5kb\nand plasmid average (%)")+
  ylab("Additional palindrome depletion in leading region (per kb)")

# Calculate actual GC and actual palindrome density
# for a given x-value. I think 5kb is appropriate for leading region to 
# get rid of start effects. 
gc_data_frame = function(position_on_plasmid){
  d.reduced %>% 
  group_by(plasmid, PTU) %>%
  mutate(GC=GC_n[x==position_on_plasmid]/5000 * 100,
         diff.GC=(GC_n[x==position_on_plasmid]-mean(GC_n))/5000 * 100,
         density=palindrome_density[x==position_on_plasmid]*1000,
         density.depletion=diff_density[x==position_on_plasmid]*1000) %>%
  filter(x==position_on_plasmid)
}
d.GC.palindrome = gc_data_frame(5000)
ggplot(d.GC.palindrome, aes(diff.GC, density.depletion))+
  geom_point()

d.GC.palindrome.PTU = d.GC.palindrome %>% group_by(PTU) %>%
  summarise(GC=mean(GC),
            diff.GC=mean(diff.GC),
            density=mean(density),
            density.depletion=mean(density.depletion),
            n=length(window_size))
ggplot(d.GC.palindrome, aes(diff.GC, density.depletion, fill=PTU))+
  geom_point(shape=21, colour="black")+
  facet_wrap(~PTU)
p.gc.correlation = ggplot(d.GC.palindrome, aes(diff.GC, density.depletion))+
  geom_point()+
  theme_bw()+
  xlab("Relative GC content at 5-10kb (%)")+
  ylab("Relative palindrome depletion at 5-10kb (per kb)")+
    xlim(c(-15, 10))+
  ylim(-13, 14)
p.gc.correlation.ptu =  ggplot(d.GC.palindrome.PTU, aes(diff.GC, density.depletion))+
  geom_point(aes(size=n, fill=PTU), shape=21, colour="black")+
  theme_bw()+
  ggrepel::geom_text_repel(aes(label=PTU), size=2, min.segment.length = 0)+
  guides(fill=FALSE)+
    xlab("Relative GC content (%)")+
  ylab("Relative palindrome depletion (per kb)")
pdf('../results/GC-correlation-depletion.pdf', width=8, height=4)
cowplot::plot_grid(p.gc.correlation+ggtitle("(a)"), p.gc.correlation.ptu+ggtitle("(b)"))
dev.off()
cor.test(d.GC.palindrome$diff.GC, d.GC.palindrome$density.depletion, method="spearman")
cor.test(d.GC.palindrome.PTU$diff.GC, d.GC.palindrome.PTU$density.depletion)

summary(lm(density.depletion ~ diff.GC*PTU, data=d.GC.palindrome))

### Averaging over all plasmids

@@

```{r}
overall.median = d.reduced %>% filter(x<50000) %>%
  group_by(x) %>%
  summarise(y=median(diff_density),
            n=length(diff_density))
p.overall.average = ggplot(overall.median, aes(x, y*1000))+
  geom_line()+
  scale_y_continuous(breaks=seq(-10, 0, 2), limits = c(-9, 0))+
  theme_bw()+
  theme(panel.grid = element_blank())+
  geom_hline(yintercept = 0, linetype='dashed')+
  ylab("Palindrome depletion (per kb)")+
    scale_x_continuous(labels = function(x) x / 1000) +
  xlab("Position on plasmid (kb)")+
  theme(axis.text=element_text(colour="black"))
p.overall.average
# Save
ggsave(p.overall.average, file='../results/6mer-deviation-overall.pdf', 
       width=3, height=3)
```
Also do this for GC content.

```{r}
overall.median.GC = d.reduced %>% filter(x<50000) %>%
  group_by(x) %>%
  summarise(y=mean(GC_n),
            n=length(GC_n))
p.overall.average.GC = ggplot(overall.median.GC, aes(x, y/5000 * 100))+
  geom_line()+
  theme_bw()+
  theme(panel.grid = element_blank())+
  geom_hline(yintercept = 50, linetype='dashed')+
  ylab("GC content (per kb)")+
    scale_x_continuous(labels = function(x) x / 1000) +
  xlab("Position on plasmid (kb)")+
  theme(axis.text=element_text(colour="black"))
p.overall.average.GC
# Save
ggsave(p.overall.average.GC, file='../results/GC-mean-overall.pdf', 
       width=5, height=2)
```
